Nonparametric Goodness-of-Fit Tests for Discrete Null Distributions

### 2010-11-19 rjournal/RJtemplate.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/ks-test@32 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

This paper presents a revision of R's
\code{ks.test()} function and a new \code{cvm.test()} function to
fill this void for researchers and practitioners in the R environment. 
We first present overviews of the theory and general implementation of the
discrete Kolmogorov-Smirnov and Cram\'{e}r-von Mises tests.  We discuss
the particular implementation of tests in R and provide examples.  We
conclude with a short discussion, including the state of existing continuous
and two-sample Cram\'{e}r-von Mises testing in R.

@@

The extension of the Kolmogorov-Smirnov test to non-continuous 
null distributions is not straightforward. The formula of
the test statistic $D$ remains unchanged, but its distribution
is much more difficult to obtain; unlike the 
continuous case, its distribution depends on the null model.
It has been known since at least the 1950's that using the tables associated
with continuous null distributions results in conservative p-values
when the null distribution is discontinuous 
(see *** or *** or \citet{stephens1974}, for example).  In the early 1970's, 
\citet{Conover1972} developed the method implemented here
for computing exact p-values
in the case of discrete null distributions.

@@

The form of the test statistic is the same as in the
continuous case; it would seem that no additional work would be required
for the implemenetation, but this is not the case.
Consider two non-decreasing functions $f$ and $g$, where the function $f$ is a step function with jumps on the set $\{x_1, \ldots x_N \}$ and $g$
is continuous (the classical Kolmogorov-Smirnov situation).
In order to determine the supremum of the difference between
these two functions, notice that **** FORMAT THESE AT END ****
\begin{align}
&\sup_x \left| f(x)- g(x) \right|  \nonumber  \\
       &= \max_i \left( \max\left\{ \left|g(x_i) - f(x_i) \right|, \lim_{x \rightarrow x_i} \left| g(x) - f(x_{i-1}) \right| \right\} \right) \label{SUP1}\\
&=  \max_i \bigg(\max \bigg\{ \left|g(x_i) - f(x_i) \right|, \left| g(x_i) - f(x_{i-1}) \right| \bigg\} \bigg). \label{SUP2}
\end{align}
Computing the maximum over these $2N$
values (with $f$ equal to
$F_{data}(x)$ and $g$ equal to $F_0(x)$ as defined above) is clearly the 
most efficient way to compute the Kolmogorov-Smirnov test statistic for
a continuous null distribution. When the function $g$ is not
continuous, however, equality (\ref{SUP2}) does not hold in general because 
we cannot replace $\lim_{x\rightarrow x_i} g(x)$ with the value $g(x_i)$.

If it is known that $g$ is a step function, it follows that
for some small $\epsilon$,
\begin{align}
\sup_x &\left| f(x)- g(x) \right| =  \nonumber \\
        &                    \max_i \{ \left|g(x_i) - f(x_i) \right|, \left| g(x_i - \epsilon) - f(x_{i-1}) \right| \} \label{epsilon}
\end{align}
where the discontinuities in $g$ are more than some distance $\epsilon$ apart. 
This, however, requires knowledge that $g$ is a step function as well as of
the nature of its support (or break-points).
%
%*** Hmm... do we care about time? ***  When the break-points of $g$ are
%unknown, or when $g$ is neither a step function nor a continuous distribution
%function, computing the supremum requires taking the numerical limit in (1***?). 
%This is computational involved and risks problems with numerical tolerances.
%
We implement the Kolmogorov-Smirnov test statistic for discrete null
distributions by forcing the user to completely specify the null distribution.

%**** Hmmm... It is often the case that the test is used inside of a
%long simulation, 
%and the added computational time would likely be prohibitive.

@@

%** code hasn't previously existed?  I deleted this stuff, not sure if
%it fits here.  Is it obvious (since we bothered to do this in the first
%place)?  Or move it?

%Code for this procedure in the R language, or in any other open source options, did not previously
%exist and is included in the new function \pkg{ks.test}.

@@

While the Kolmogorov-Smirnov test may be the most popular of
the nonparametric goodness-of-fit tests, there is another family of 
tests which has been shown to be more powerful against a large class
of alternatives hypotheses. 
The original test was developed by
Harald Cram\'{e}r and Richard von Mises \citep{cramer1928, vonmises1928} and further adapted by \cite{anderson1952}, and  \cite{Watson1961}. 
The original test statistic, $W^2$, Anderson's $A^2$, and Watson's
$U^2$ are:
\begin{align}
W^2 &= n \cdot \int_{-\infty}^{\infty} \left[ F_{data}(x)- F_{0}(x) \right]^2 dF_0(x) \label{W2} \\
A^2 &= n \cdot \int_{-\infty}^{\infty} \frac{\left[F_{data}(x)- F_{0}(x) \right]^2}{F_0(x) -F_0(x)^2} dF_0(x) \label{A2} \\
U^2 &= n \cdot \int_{-\infty}^{\infty} \left[ F_{data}(x)- F_{0}(x) - W^2 \right]^2 dF_0(x) \label{U2}
\end{align}
As with the original Kolmogorov-Smirnov test statistic, these all have 
null distributions of their test statistics which are independent of the
specified null models.

The $W^2$
statistic was the original test statistic; it has the benefit of having
the simplest form.
The $A^2$ statistic was developed by
Anderson in the process of generalizing the test for the two-sample case.
Watson's $U^2$ statistic was developed for distributions
which are cyclic (with an ordering to the support but
no natural starting point), as it is invariant to cyclic reordering of
the support. 
For example, a distribution on the months of the year could be considered
cyclic.

It has been shown that these tests can be more powerful than Kolmogorov-Smirnov tests to certain deviations from the hypothesized distribution. 
As they all involve
integration over the whole range of data, rather than use of a supremum.
Thus, they are best-suited for situations where the
true alternative distribution deviates a little over the whole support
rather than having large deviations over a small section of the support. For
a complete analysis of the relative powers of these tests, see 
\cite{stephens1974}.

@@

Generalizations of the Cram\'{e}r-von Mises tests to discrete
distributions were developed in \cite{choulakian1994}. As with the
Kolmogorov-Smirnov test, the forms of the test statistics are unchanged, 
and the null distributions of the test statistics are again hypothesis-dependent. The methods do not offer finite-sample results, but rather 
show that the asymptotic distributions of the test statistics under the null hypothesis each involve consideration of a weighted sum of independent
chi-squared variables (with the weights depending on the particular
null distribution). 
We implement the asymptotic distributions here; \cite{stephens1974}
showed that these approximations are conservative 
and asymptotically equivalent to the true null distributions.

@@

There is no analytic solution to the integral in (\ref{distQ}), 
so the integration is
accomplished numerically. This seems fine in most situations we considered,
but numerical issues appear in the regime of large test statistics $x$
(or, equivalently, small p-values).
The function $\theta(\cdot, x)$ is linear in $x$; as the test statistic 
grows the corresponding periodicity of the integrand decreases and
the approximation becomes unstable. 
%This is further magnified in situations where the 
%$p$-values should be very small, where tiny fluctuations (undetectable
%elsewhere) become quite prominent. 
We resolve this problem by using a simple conservative approximation to
avoid the numerical instability. Consider the
following inequality:
\begin{align}
\mathbb{P} \left(\sum_{i=1}^{p} \lambda_i \chi^2_1 \geq x \right) &\leq \mathbb{P} \left( \lambda_{max} \sum_{i=1}^{p} \chi^2_1 \geq x \right) \label{ineq1} \\
&= \mathbb{P} \left(\chi^2_p \geq \frac{x}{p \, \lambda_{max}} \right)
\label{ineq2}
\end{align}
The values for the weighted sum can be bounded using a simple transformation
and a chi-squared distribution of a higher degree of freedom. 
The original formulation is preferable for most p-values, while
this approximation is used for smaller p-values (smaller than 0.001,
based on our observations of the numerical instability of the original
formulation).

Figure \ref{cvmissues} shows the non-monotonicity of the numerical
evaluation of equation (\ref{distQ}) for the null distribution which is
uniform on the set $\{1,2,3\}$, as well as the behavior of the
conservative upper bound given in equation (\ref{ineq2}).

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{fig1.pdf}
\end{center}
\caption{Plot of calculated p-values for given test statistics
using numerical integration
(pink) compared to the conservative chi-squared bound (dashed blue). 
The null distribution is uniform on the set $\{1,2,3\}$ in this example.
The instability in the calculated p-values is a result of numerical
instabilities.}
\label{cvmissues}
\end{figure}

@@

Functions \code{ks.test()} and \code{cvm.test()} are provided for
convenience in packages \pkg{ks.test} and \pkg{cvm.test}, respectively.
Function \code{ks.test()} offers a revision of
R's Kolmogorov-Smirnov function \code{ks.test()} from base
package \pkg{stats}, while \code{cvm.test()} is a new
function for Cram\'{e}r-von Mises tests.
Both are available from the authors
or from R-Forge (\url{https://r-forge.r-project.org/R/?group_id=802});
they will be proposed for inclusion in \pkg{stats} in late 2010.

>>>>

This paper presents a revision of R's
\code{ks.test()} function and a new \code{cvm.test()} function to
fill this void for researchers and practitioners in the R environment
\citep{R2010}. 
We first present overviews of the theory and general implementation of the
discrete Kolmogorov-Smirnov and Cram\'{e}r-von Mises tests.  We discuss
the particular implementation of the tests in R and provide examples.  We
conclude with a short discussion, including the state of existing continuous
and two-sample Cram\'{e}r-von Mises testing in R.

@@

The extension of the Kolmogorov-Smirnov test to non-continuous 
null distributions is not straightforward. The formula of
the test statistic $D$ remains unchanged, but its distribution
is much more difficult to obtain; unlike the 
continuous case, its distribution depends on the null model.
It has been known since at least the 1950's that using the tables associated
with continuous null distributions results in conservative p-values
when the null distribution is discontinuous 
(see *** or *** or \citet{stephens1974}, for example).  In the early 1970's, 
\citet{Conover1972} developed the method implemented here
for computing p-values
in the case of discrete null distributions.

@@

The form of the test statistic is the same as in the
continuous case; it would seem that no additional work would be required
for the implemenetation, but this is not the case.
Consider two non-decreasing functions $f$ and $g$, where the function $f$ is a step function with jumps on the set $\{x_1, \ldots x_N \}$ and $g$
is continuous (the classical Kolmogorov-Smirnov situation).
In order to determine the supremum of the difference between
these two functions, notice that
\begin{align}
\sup_x \left| f(x) - g(x) \right| &  \nonumber  \\
       =  \max_i \bigg[ \max\bigg(  & \left|g(x_i) - f(x_i) \right|, \nonumber \\
& \lim_{x \rightarrow x_i} \left| g(x) - f(x_{i-1})
   \right| \bigg) \bigg] \label{SUP1}\\
=  \max_i \bigg[\max \bigg( & \left|g(x_i) - f(x_i) \right|, \nonumber \\
& \left| g(x_i) - f(x_{i-1}) \right| \bigg) \bigg]. \label{SUP2}
\end{align}
Computing the maximum over these $2N$
values (with $f$ equal to
$F_{data}(x)$ and $g$ equal to $F_0(x)$ as defined above) is clearly the 
most efficient way to compute the Kolmogorov-Smirnov test statistic for
a continuous null distribution. When the function $g$ is not
continuous, however, equality (\ref{SUP2}) does not hold in general because 
we cannot replace $\lim_{x\rightarrow x_i} g(x)$ with the value $g(x_i)$.

If it is known that $g$ is a step function, it follows that
for some small $\epsilon$,
\begin{align}
\sup_x &\left| f(x)- g(x) \right| =  \nonumber \\
        & \max_i \left( \left|g(x_i) - f(x_i) \right|, 
    \left| g(x_i - \epsilon) - f(x_{i-1}) \right| \right) \label{epsilon}
\end{align}
where the discontinuities in $g$ are more than some distance $\epsilon$ apart. 
This, however, requires knowledge that $g$ is a step function as well as of
the nature of its support (or break-points).  As a result,
we implement the Kolmogorov-Smirnov test statistic for discrete null
distributions by requiring the complete specification of the null distribution.

@@

@@

While the Kolmogorov-Smirnov test may be the most popular of
the nonparametric goodness-of-fit tests, Cram\'{e}r-von Mises
tests have been shown to be more powerful against a large class
of alternatives hypotheses. 
The original test was developed by
Harald Cram\'{e}r and Richard von Mises \citep{cramer1928, vonmises1928} and further adapted by \cite{anderson1952}, and  \cite{Watson1961}. 
The original test statistic, $W^2$, Anderson's $A^2$, and Watson's
$U^2$ are:
\begin{align}
W^2 &= n \cdot \int_{-\infty}^{\infty} \left[ F_{data}(x)- F_{0}(x) \right]^2 dF_0(x) \label{W2} \\
A^2 &= n \cdot \int_{-\infty}^{\infty} \frac{\left[F_{data}(x)- F_{0}(x) \right]^2}{F_0(x) -F_0(x)^2} dF_0(x) \label{A2} \\
U^2 &= n \cdot \int_{-\infty}^{\infty} \left[ F_{data}(x)- F_{0}(x) - W^2 \right]^2 dF_0(x) \label{U2}
\end{align}
As with the original Kolmogorov-Smirnov test statistic, these all have 
null distributions of their test statistics which are independent of the
specified null models.

The $W^2$ statistic was the original test statistic.
The $A^2$ statistic was developed by
Anderson in the process of generalizing the test for the two-sample case.
Watson's $U^2$ statistic was developed for distributions
which are cyclic (with an ordering to the support but
no natural starting point); it is invariant to cyclic reordering of
the support.   For example, a distribution on the months of the year
could be considered cyclic.

It has been shown that these tests can be more powerful than Kolmogorov-Smirnov tests to certain deviations from the hypothesized distribution. 
They all involve integration over the whole range of data, rather than
use of a supremum, so they are best-suited for situations where the
true alternative distribution deviates a little over the whole support
rather than having large deviations over a small section of the support. 
\cite{stephens1974} offers a complete analysis of the relative powers
of these tests.

@@

Generalizations of the Cram\'{e}r-von Mises tests to discrete
distributions were developed in \cite{choulakian1994}. As with the
Kolmogorov-Smirnov test, the forms of the test statistics are unchanged, 
and the null distributions of the test statistics are again hypothesis-dependent. \cite{choulakian1994} does not offer finite-sample results, but rather 
shows that the asymptotic distributions of the test statistics under the null hypothesis each involve consideration of a weighted sum of independent
chi-squared variables (with the weights depending on the particular
null distribution). 
We implement the asymptotic distributions here; \cite{stephens1974}
showed that these approximations are conservative 
and asymptotically equivalent to the true null distributions.

**** Why would Stephens (1974) have this, when they were developed in 1994? ****

@@

There is no analytic solution to the integral in (\ref{distQ}), 
so the integration is
accomplished numerically. This seems fine in most situations we considered,
but numerical issues appear in the regime of large test statistics $x$
(or, equivalently, small p-values).
The function $\theta(\cdot, x)$ is linear in $x$; as the test statistic 
grows the corresponding periodicity of the integrand decreases and
the approximation becomes unstable. 
As an example of this numerical instability, the red plotted in
Figure \ref{cvmissues} shows the non-monotonicity of the numerical
evaluation of equation (\ref{distQ}) for a null distribution that is
uniform on the set $\{1,2,3\}$.

We resolve this problem by using a simple conservative approximation to
avoid the numerical instability. Consider the
following inequality:
\begin{align}
\mathbb{P} \left(\sum_{i=1}^{p} \lambda_i \chi^2_1 \geq x \right) &\leq \mathbb{P} \left( \lambda_{max} \sum_{i=1}^{p} \chi^2_1 \geq x \right) \label{ineq1} \\
&= \mathbb{P} \left(\chi^2_p \geq \frac{x}{p \, \lambda_{max}} \right)
\label{ineq2}
\end{align}
The values for the weighted sum can be bounded using a simple transformation
and a chi-squared distribution of a higher degree of freedom. 
The original formulation is preferable for most p-values, while
this upper bound is used for smaller p-values (smaller than 0.001,
based on our observations of the numerical instability of the original
formulation).  Although it would be preferable to determine the use of
the bound based on values of the test statistic rather than the p-value,
the range of ``extreme'' values of the test statistic varies with the
hypothesized distribution.

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{fig1.pdf}
\end{center}
\caption{Plot of calculated p-values for given test statistics
using numerical integration
(red) compared to the conservative chi-squared bound (dashed blue). 
The null distribution is uniform on the set $\{1,2,3\}$ in this example.
The instability in the calculated p-values is a result of numerical
instabilities.}
\label{cvmissues}
\end{figure}

@@

Functions \code{ks.test()} and \code{cvm.test()} are provided for
convenience in packages \pkg{ks.test} and \pkg{cvm.test}, respectively.
Function \code{ks.test()} offers a revision of
R's Kolmogorov-Smirnov function \code{ks.test()} from base
package \pkg{stats}, while \code{cvm.test()} is a new
function for Cram\'{e}r-von Mises tests.
Both are available from the authors
or from R-Forge (\url{https://r-forge.r-project.org/R/?group_id=802});
\code{ks.test()} will be proposed for inclusion in \pkg{stats} in late 2010.
